### 2017-07-03 isolation.tex
new opsem

Databases implement isolation levels through a combination of various
concurrency control mechanisms, which are absent from
Fig.~\ref{fig:txnimp}. In this section, we first axiomatize the
semantics of various concurrency control constructs, and then define
the precise semantics of the isolation levels as they are implemented
on Postgres and MySQL datbases.
% Any additional concurrency control provided by the database
% implementation, beyond that which is already incuded in the
% isolation specification, also needs to be axiomatized in $\I$. We
% first describe such axiomatizations, and then present the
% specifications of the standard isolation levels as implemented by
% popular off-the-shelf databases.

\textbf{Unique Ids}. Most relational databases associate every record
in a table with a unique identifier. Often, auto-generated unique
identifiers are part of the schema itself (via SQL's \C{UNIQUE} and
\C{AUTO\_INCREMENT} keywords), but if not, database adds one (e.g.,
MySQL's InnoDB engine automatically adds a 6-byte identifier if none
exists). Enforcing global uniqueness of identifiers requires exclusive
locks, which are absent from Fig.~\ref{fig:txnimp}. The
\rulelabel{E-Insert} rule assigns an id to the newly added record
after checking that it is not already present in $\stg$ and $\stl$.
But the new record is only added to $\stl$, which is invisible outside
the transaction. Thus, the uniqueness check for the same id also
passes in a concurrent transaction, resulting in duplicate ids
after both commits. To prevent this, we need the following proposition
in $\I$ of every transaction, regardless of whether it is executing
or committing:
\begin{smathpar}
\begin{array}{lcl}
  \I_{id}(\stl,\stg,\stg') & = & \forall(r\in\stl).~
      r.\idf\notin \dom(\stg) \Rightarrow r.\idf\notin \dom(\stg').
\end{array}
\end{smathpar}
$\I_{id}$ makes sure that a globally unique id generated during a
transaction's execution remains globally unique until it commits. An
interference from a concurrent transaction that adds the same id to
$\stg$ is prohibited.

\textbf{Write Locks}. Transactions in MySQL and Postgres (effectively)
obtain exclusive locks on records they update, and do not release them
until they commit, thus preventing write-write (ww) conflicts. To
obtain this behavior on the abstract machine of Fig.~\ref{fig:txnimp},
the following proposition needs to be in $\I$ for every transaction:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ww}(\stl,\stg,\stg') & = & \forall(r'\in\stl)(r \in \stg).~
      r.\idf = r'.\idf  \Rightarrow r\in\stg'.
\end{array}
\end{smathpar}
That is, given a record $r'\in\stl$, if there exists an $r\in\stg$
with the same id (i.e., $r'$ is an updated version of $r$), then $r$
must be present unmodified in $\stg'$. This prevents a concurrent
transaction from changing $r$, thus simulating the behavior of an
exclusive lock. There is however a caveat: if we assume extensional
equality over records, a write by a concurrent transaction that
doesn't change $r$'s contents is still allowed. While this in itself
is not a problem, the concurrent transaction may modify other records,
which then become visible in the current transaction. A write lock
prevents this behavior, whereas our axiomatization ($\I_{ww}$) allows
it. An easy fix for this is to add version timestamps to records that
effectively intensionalizes equality. Nonetheless, imprecise
axiomatization of write locks hasn't been a problem in practice.

\textbf{Snapshots} Both MySQL and Postgres implement isolation levels
that execute transactions against a static snapshot of the database.
The semantics of a snapshot be axiomatized as following:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ss}(\stl,\stg,\stg') & = & \stg' = \stg.
\end{array}
\end{smathpar}
% The snapshot semantics however may not be applicable throughout the
% lifetime of the transaction, like in the case of Repeatable Read (RR).

@@

%\usepackage[table,xcdraw]{xcolor}
\begin{table}[]
\centering
\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|}
\cline{2-13}
                                & \multicolumn{6}{c|}{{\color[HTML]{333333} Postgres}}                         & \multicolumn{6}{c|}{MySql}                                                   \\ \cline{2-13} 
                                & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} \\ \cline{2-13} 
                                & T          & F          & T          & F          & T           & F          & T          & F          & T          & F          & T           & F          \\ \hline
\multicolumn{1}{|l|}{$\I_{ss}$} &            &            & \checkmark &            & \checkmark  & \checkmark &            &            & \checkmark &            & \checkmark  & \checkmark \\ \hline
\multicolumn{1}{|l|}{$\I_{rb}$} &            &            &            & \checkmark &             &            &            &            &            &            &             &            \\ \hline
\multicolumn{1}{|l|}{$\I_{ro}$} &            &            &            &            &             &            &            &            & \checkmark & \checkmark &             &            \\ \hline
\multicolumn{1}{|l|}{$\I_{id}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
\multicolumn{1}{|l|}{$\I_{ww}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
\end{tabular}

\caption{The semantics of isolation levels on Postgres and MySQL}
\label{fig:iso-table}
\end{table}

With help of the above axiomatizations, we can now define the precise
semantics of various isolation levels on Postgres and MySQL. The table
in Table~\ref{fig:iso-table} captures the summary.

>>>>

Databases implement isolation levels through a combination of various
concurrency control mechanisms, which are absent from
Fig.~\ref{fig:txnimp}. In this section, we first axiomatize the
semantics of various concurrency control constructs, and then define
the precise semantics of the isolation levels as they are implemented
on two real-world databases -  PostgreSQL and MySQL.
% Any additional concurrency control provided by the database
% implementation, beyond that which is already incuded in the
% isolation specification, also needs to be axiomatized in $\I$. We
% first describe such axiomatizations, and then present the
% specifications of the standard isolation levels as implemented by
% popular off-the-shelf databases.

\textbf{Unique Ids}. Most relational databases associate every record
in a table with a unique identifier. Often, auto-generated unique
identifiers are part of the schema itself (via SQL's \C{UNIQUE} and
\C{AUTO\_INCREMENT} keywords), but if not, database adds one (e.g.,
MySQL's InnoDB engine automatically adds a 6-byte identifier if none
exists). Enforcing global uniqueness of identifiers requires exclusive
locks, which are absent from Fig.~\ref{fig:txnimp}. The
\rulelabel{E-Insert} rule assigns an id to the newly added record
after checking that it is not already present in $\stg$ and $\stl$.
But the new record is only added to $\stl$, which is invisible outside
the transaction. Thus, it is possible that the uniqueness check for the same id may
also succeed in a concurrent transaction, resulting in duplicate ids
after both commits. To prevent this, we need the following proposition
in $\I$ of every transaction, regardless of whether it is executing
or committing:
\begin{smathpar}
\begin{array}{lcl}
  \I_{id}(\stl,\stg,\stg') & = & \forall(r\in\stl).~
      r.\idf\notin \dom(\stg) \Rightarrow r.\idf\notin \dom(\stg').
\end{array}
\end{smathpar}
$\I_{id}$ ensures that a globally unique id generated during a
transaction's execution remains globally unique until it commits, prohibiting
interference from a concurrent transaction that adds the same id.

\textbf{Write Locks}. Transactions in MySQL and PostgreSQL (effectively)
obtain exclusive locks on records they update, and do not release them
until they commit, thus preventing write-write (ww) conflicts. To
obtain this behavior on the abstract machine of Fig.~\ref{fig:txnimp},
the following proposition must hold for every transaction:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ww}(\stl,\stg,\stg') & = & \forall(r'\in\stl)(r \in \stg).~
      r.\idf = r'.\idf  \Rightarrow r\in\stg'.
\end{array}
\end{smathpar}
That is, given a record $r'\in\stl$, if there exists an $r\in\stg$
with the same id (i.e., $r'$ is an updated version of $r$), then $r$
must be present unmodified in $\stg'$. This prevents a concurrent
transaction from changing $r$, thus simulating the behavior of an
exclusive lock. There is however a caveat: if we assume extensional
equality over records, a write by a concurrent transaction that
doesn't change $r$'s contents is still allowed. While this in itself
is not a problem, the concurrent transaction may modify other records,
which then become visible in the current transaction. A write lock
prevents this behavior, whereas our axiomatization ($\I_{ww}$) allows
it. An easy fix for this is to add version timestamps to records that
effectively intensionalizes equality. Nonetheless, imprecise
axiomatization of write locks hasn't been a problem in practice.

\textbf{Snapshots} Both MySQL and PostgresSQL implement isolation levels
that execute transactions against a static snapshot of the database.
The semantics of a snapshot be axiomatized as following:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ss}(\stl,\stg,\stg') & = & \stg' = \stg.
\end{array}
\end{smathpar}
% The snapshot semantics however may not be applicable throughout the
% lifetime of the transaction, like in the case of Repeatable Read (RR).

@@

%\usepackage[table,xcdraw]{xcolor}
\begin{table}[]
\centering
\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|}
\cline{2-13}
                                & \multicolumn{6}{c|}{{\color[HTML]{333333} PostgreSQL}}                         & \multicolumn{6}{c|}{MySql}                                                   \\ \cline{2-13} 
                                & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} \\ \cline{2-13} 
                                & T          & F          & T          & F          & T           & F          & T          & F          & T          & F          & T           & F          \\ \hline
\multicolumn{1}{|l|}{$\I_{ss}$} &            &            & \checkmark &            & \checkmark  & \checkmark &            &            & \checkmark &            & \checkmark  & \checkmark \\ \hline
\multicolumn{1}{|l|}{$\I_{rb}$} &            &            &            & \checkmark &             &            &            &            &            &            &             &            \\ \hline
\multicolumn{1}{|l|}{$\I_{ro}$} &            &            &            &            &             &            &            &            & \checkmark & \checkmark &             &            \\ \hline
\multicolumn{1}{|l|}{$\I_{id}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
\multicolumn{1}{|l|}{$\I_{ww}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
\end{tabular}

\caption{The semantics of isolation levels on PostgreSQL and MySQL}
\label{fig:iso-table}
\end{table}

With help of the above axiomatizations, we can now define the precise
semantics of various isolation levels on PostgreSQL and MySQL. The table
in Table~\ref{fig:iso-table} captures the summary.

### 2017-07-03 motivation.tex
new opsem

\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-postgres}
\caption{Database state transitions corresponding to an execution of
  an RR transaction $T_1$ on PostgresSQL. $T_2$ and $T_3$ are concurrent
  transactions. The database state has three data items, $a$, $b$ and $c$. }
\label{fig:rr-postgres}
\end{figure}
\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-abstract}
  \caption{An abstract execution that includes the concrete execution
  shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
  admits fewer interleavings, yet results in the same post-state.  }
\label{fig:rr-abstract}
\end{figure}

@@

We can axiomatize this operational description by observing that (a)
due to the version check, the current transaction cannot update a data
item that was already updated by a concurrent transaction, and (b) due
to the use of exclusive write locks, a data item updated by the
current transaction cannot be overwritten by a concurrent transaction.
If $\Delta$ is the state of the database when an RR transaction
finishes, and $\Delta_c$ is the state visible to the transaction at
the point of commit, we know the transition from $\Delta$ to
$\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
effects from any concurrent transactions that write to the same data
items as the current RR transaction.  Similarly, if $\delta$ denotes
a local log that relates transaction variables being written with their updated values,
then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
\Delta(x)$. To summarize, the operational semantics of PostgresSQL's RR
implementation can be captured as an axiomatization over transitions of
the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an RR transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

>>>>

\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-postgres}
\caption{Database state transitions corresponding to an execution of
  an RR transaction $T_1$ on PostgreSQL. $T_2$ and $T_3$ are concurrent
  transactions. The database state has three data items, $a$, $b$ and $c$. }
\label{fig:rr-postgres}
\end{figure}
\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-abstract}
  \caption{An abstract execution that includes the concrete execution
  shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
  admits fewer interleavings, yet results in the same post-state.  }
\label{fig:rr-abstract}
\end{figure}

@@

We can axiomatize this operational description by observing that (a)
due to the version check, the current transaction cannot update a data
item that was already updated by a concurrent transaction, and (b) due
to the use of exclusive write locks, a data item updated by the
current transaction cannot be overwritten by a concurrent transaction.
If $\Delta$ is the state of the database when an RR transaction
finishes, and $\Delta_c$ is the state visible to the transaction at
the point of commit, we know the transition from $\Delta$ to
$\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
effects from any concurrent transactions that write to the same data
items as the current RR transaction.  Similarly, if $\delta$ denotes
a local log that relates transaction variables being written with their updated values,
then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
\Delta(x)$. To summarize, the operational semantics of PostgreSQL's RR
implementation can be captured as an axiomatization over transitions of
the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an RR transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

### 2017-07-03 opsem.tex
new opsem

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), Integer and boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\{\bar{f}=\bar{e}\}$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$), and
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). The $\I$ in \C{TXN} block syntax is the
transaction's isolation specification, which will be explained later.
Certain terms that only appear at run-time are also present in $c$.
These include a \C{TXN} block tagged with sets ($\stl$ and $\stg$) of
records that assume special meaning in operational semantics, and a
\C{FOREACH} loop that keeps track of the set ($s_1$) of items already
iterated, and the set ($s_2$) of items yet to be iterated. Note that
even the surface-level \C{FOREACH} shown here is a little different
from the one used in previous sections; its higher-order argument has
two arguments, $y$ and $z$, which are invoked (during the reduction)
with the set of already-iterated items, and the current item,
respectively. This form of \C{FOREACH} lends itself to inductive
reasoning, facilitating inductive proofs (Sec.~\ref{sec:reasoning}).

% also We let $T_i$ for $i \in \mathbb{N}$
% range over transaction identifiers. When it is evident we are
% referring to a transaction, we use the number $i$ instead of $T_i$ for
% identification (\eg in $\C{txn}\langle i \rangle$). Like variables,
% transaction identifiers are globally accessible. For notational
% convenience, we let $t$ range over both expressions and commands.

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local database ($\stl$), or the global database
($\stg$). Database ($\stg$) is a modeled as a set of records of a
pre-defined type, i.e., they all belong to a single table.
Generalization to multiple tables is straightforward. Records in
$\stg$ are uniquely identifiable through their $\idf$ field, which is
auto-generated and does not belong to the surface language, i.e.,
$\C{id}\notin f$. For a set $S$ of records, we define $\dom(S)$ as the
set of unique ids of all the records in $S$. Thus $|\dom(\stg)| =
|\stg|$. During its execution, a transaction may write to multiple
records in $\stg$. Atomicity dictates that such writes shouldn't be
available in $\stg$ until the transaction commits. We therefore
associate each transaction with a local database ($\stl$) that stores
\emph{only} the uncommitted records\footnote{While SQL's \C{UPDATE}
admits writes at the granularity of record fields, databases, in
reality, enforce record-level locking, allowing us to think of
``uncommitted writes'' as ``uncommitted records''. }. Uncommitted
records include deleted records, which are identified by a hidden
$\delf$ field set to \C{true}. When the transaction commits, its local
database is atomically \emph{flushed} to the global database,
committing the uncommitted records. The flush operation ($\gg$) is defined as
following:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\gg\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and not been updated in $\stl$, i.e., $r.\idf \notin
\dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new record,
or an updated version of an old record, but the updatation is not a
deletion ($\neg r.\delf$). Thus, flush defines the result of
atomically applying transaction's local writes to the global database.
Besides the commit, flush also helps a transaction read its own
writes. Intuitively, the result of a read operation inside a
transaction must be computed on the database resulting from flushing
the current local state ($\stl$) to the global state ($\stg$). The
abstract machine of Fig.~\ref{fig:txnimp}, however, doesn't let a
transaction read its own writes. This reduces verbosity and simplifies
semantics, without losing any generality; substituting $\stl\gg\stg$
for $\stg$ at select places in reduction rules recovers the real

Small-step semantics is stratified into a transaction-local reduction
relation, and a top-level reduction relation. Transaction-local
relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
small-step reduction for a command inside a transaction, when the
database state is $\stg$; the command $c$ reduces to $c'$, while
updating the transaction-local database $\stl$ to $\stl'$. The
definition assumes a meta function $\eval$ that evaluates expressions
with no free variables to values. The reduction relation for SQL
statements is defined straightforwardly.  \C{INSERT} adds a new record
to $\stl$ after adding a unique identifier (more discussion on
uniqueness later). \C{DELETE} finds the records in $\stg$ that match
the search criteria defined by its boolean function argument, and adds
the records to $\stl$ after marking them for deletion. \C{SELECT}
bounds the name introduced by \C{LET} to the set of records from
$\stg$ that match the search criteria, and then executes the bound
command $c$. \C{UPDATE} uses its first function argument to compute
the updated version of the records that match the search criteria
defined by its second function argument. Updated records are added to
$\stl$.

The reduction of \C{FOREACH} starts by first converting it to its
run-time form that also keeps track of the iterated items ($s_1$),
besides the yet-to-be-iterated items ($s_2$). Initially, $s_1$ is
empty. As the elements are iterated, they are removed from $s_2$ and
added to $s_1$ . Iteration involves invoking the higher-order function
with $s_1$ and the current element $x$ (note: $\uplus$ in $\{x\}
\uplus s_2$ denotes a disjoint union). The reduction ends when the
$s_2$ becomes empty. The reduction rules for conditionals, \C{LET}
binders, and sequences are straightforward, hence ommitted.

The top-level reduction relation defines small-step semantics of
transactions, and their parallel composition. A transaction always
comes tagged with an isolation specification $\I$ that dictates the
timing and nature of iterferences that the transaction can witness.
Formally, $\I$ is a function that maps a boolean ($\mathbb{B}$) and a
triplet of database states (three sets of records:
$\Pow{r}\times\Pow{r}\times\Pow{r}$) to a logical proposition
($\Prop$). The abstract machine calls $\I$ while reducing
$\ctxn{i}{\I}{c}$. The boolean argument identifies if the transaction
is still executing (\C{true}), or if it has finished execution and
ready to commit (\C{false}).  Among the triplet of sets (the second
argument), first is the (current) transaction-local database state
($\stl$), second is the state ($\stg$) of the global database when the
transaction last took a step, and third is the current state ($\stg'$)
of the global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from a concurrent transaction, and the $\I$ decides if
this interference is allowed or not, considering the local database
state ($\stl$). For instance, as described in
(Sec.~\ref{sec:motivation}), an RR transaction on Postgres defines
$\I$ as following:
\begin{smathpar}
\begin{array}{lcl}
\I\,\C{true}\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I\,\C{false}\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~
  r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}

The reduction of a $\ctxn{i}{\I}{c}$ begins by first converting it to
its run-time form $\ctxn{i}{\I,\stl,\stg}{c}$, where $\stl =
\emptyset$, and $\stg$ the current (global) database. The top-level
reduction rules in Fig.~\ref{fig:txnimp} assume this form. Rule
\rulelabel{E-Txn} reduces $\ctxn{i}{\I,\stl,\stg}{c}$ under a database
state ($\stg'$), only if the isolation specification ($\I$) allows the
interference between $\stg$ and $\stg'$ during the execution. Rule
\rulelabel{E-Commit} commits the transaction
$\ctxn{i}{\I,\stl,\stg}{c}$ by flushing its uncommitted records to the
database. This is done only if the interference between $\stg$ and
$\stg'$ is allowed during the commit by the isolation specification.

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), Integer and boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\{\bar{f}=\bar{e}\}$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$), and
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). The $\I$ in the \C{TXN} block syntax is the
transaction's isolation specification, which will be explained later.
Certain terms that only appear at run-time are also present in $c$.
These include a \R{TXN} block tagged with sets ($\stl$ and $\stg$) of
records whose meaning is explained below, and a runtime \R{FOREACH}
expression that keeps track of the set ($s_1$) of items already
iterated, and the set ($s_2$) of items yet to be iterated. Note that
even the surface-level syntax of the \C{FOREACH} command shown here is
a little different from the one used in previous sections; its
higher-order function has two arguments, $y$ and $z$, which are
invoked (during the reduction) with the set of already-iterated items,
and the current item, respectively. This form of \C{FOREACH} lends
itself to inductive reasoning that will be useful for verification
(Sec.~\ref{sec:reasoning}).

% also We let $T_i$ for $i \in \matshbb{N}$
% range over transaction identifiers. When it is evident we are
% referring to a transaction, we use the number $i$ instead of $T_i$ for
% identification (\eg in $\C{txn}\langle i \rangle$). Like variables,
% transaction identifiers are globally accessible. For notational
% convenience, we let $t$ range over both expressions and commands.

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local ($\stl$), or global ($\stg$) database, both
of which are modeled as a set of records of a pre-defined type, i.e.,
they all belong to a single table.  The generalization to multiple
tables is straightforward.  Records in $\stg$ are uniquely identifiable
through their $\idf$ field, which is auto-generated and does not
belong to the surface language, i.e., $\C{id}\notin f$. For a set $S$
of records, we define $\dom(S)$ as the set of unique ids of all 
records in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
transaction may write to multiple records in $\stg$. Atomicity
dictates that such writes should not be visible in $\stg$ until the
transaction commits. We therefore associate each transaction with a
local database ($\stl$) that stores such uncommitted
records\footnote{While SQL's \C{UPDATE} admits writes at the
  granularity of record fields, databases, in reality, enforce
  record-level locking, allowing us to think of ``uncommitted writes''
  as ``uncommitted records''. }. Uncommitted records include deleted
records, which are identified by a hidden $\delf$ field set to
\C{true}. When the transaction commits, its local database is
atomically \emph{flushed} to the global database, committing these
heretofore uncommitted records. The flush operation ($\gg$) is defined as
follows:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\gg\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
\notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
record, or an updated version of an old record, but the update is not
a deletion ($\neg r.\delf$). Thus, flush defines the result of
atomically applying a transaction's local writes to the global
database.  Besides the commit, flush also helps a transaction read its
own writes. Intuitively, the result of a read operation inside a
transaction must be computed on the database resulting from flushing
the current local state ($\stl$) to the global state ($\stg$). The
abstract machine of Fig.~\ref{fig:txnimp}, however, does not let a
transaction read its own writes. This simplifies the semantics,
without losing any generality, since substituting $\stl\gg\stg$ for
$\stg$ at select places in the reduction rules effectively allows
reads of uncommitted transaction writes to be realized.

The small-step semantics is stratified into a transaction-local reduction
relation, and a top-level reduction relation. Transaction-local
relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
small-step reduction for a command inside a transaction, when the
database state is $\stg$; the command $c$ reduces to $c'$, while
updating the transaction-local database $\stl$ to $\stl'$. The
definition assumes a meta function $\eval$ that evaluates expressions
with no free variables to values. The reduction relation for SQL
statements is defined straightforwardly.  \C{INSERT} adds a new record
to $\stl$ after adding a unique identifier. \C{DELETE} finds the records in $\stg$ that match
the search criteria defined by its boolean function argument, and adds
the records to $\stl$ after marking them for deletion. \C{SELECT}
bounds the name introduced by \C{LET} to the set of records from
$\stg$ that match the search criteria, and then executes the bound
command $c$. \C{UPDATE} uses its first function argument to compute
the updated version of the records that match the search criteria
defined by its second function argument. Updated records are added to
$\stl$.

The reduction of \C{FOREACH} starts by first converting it to its
run-time form to keep track of iterated items ($s_1$),
as well as  yet-to-be-iterated items ($s_2$). Initially, $s_1$ is
empty. As the elements are iterated, they are removed from $s_2$ and
added to $s_1$ . Iteration involves invoking its function argument
with $s_1$ and the current element $x$ (note: $\uplus$ in $\{x\}
\uplus s_2$ denotes a disjoint union). The reduction ends when 
$s_2$ becomes empty. The reduction rules for conditionals, \C{LET}
binders, and sequences are straightforward, and ommitted for brevity.

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction always
comes tagged with an isolation specification $\I_e$ or $\I_c$ that
dictates the timing and nature of interferences that a transaction can
witness, both during its execution ($\I_e$), and when it is about to
commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are functions that map
the (current) transaction-local database state ($\stl$), the state
($\stg$) of the global database when the transaction last took a step,
and the current state ($\stg'$) of the global database.  Intuitively,
$\stg'\neq\stg$ indicates an interference from another concurrent
transaction, and the predicates $\I_e$ and $\I_c$ decide if this
interference is allowed or not, taking into account the local database
state ($\stl$). For instance, as described in
(Sec.~\ref{sec:motivation}), an RR transaction on PostgreSQL defines
$\I$ as following:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}

The reduction of a $\ctxn{i}{\I_e}{c}$ begins by first converting it
to its run-time form $\ctxn{i}{\I_e,\stl,\stg}{c}$, where $\stl =
\emptyset$, and $\stg$ the current (global) database.  Rule
\rulelabel{E-Txn} reduces $\ctxn{i}{\I_e,\stl,\stg}{c}$ under a
database state ($\stg'$), only if the isolation specification ($\I$)
allows the interference between $\stg$ and $\stg'$ during that
step. Rule \rulelabel{E-Commit} commits the transaction
$\ctxn{i}{\I_c,\stl,\stg}{c}$ by flushing its uncommitted records to
the database. This is done only if the interference between $\stg$ and
$\stg'$ is allowed at the commit point by the isolation specification.
The distinction between $\I_e$ and $\I_c$ allows us to model realistic
database features such as snapshot isolation that isolates the effects
a transaction can witness during its execution but exposes
interferences from concurrent transactions at the commit point.
